Coalescent inference of mutation spectrum histories from sample frequency spectra

### 2019-08-17 manuscript/main.tex
fix base case for W'
piecewise constant discretization

\section*{Introduction}\label{sec:intro}

@@

\section*{Model}\label{sec:model}

@@

Let rv $M_{\delta x, b}$ denote the event that a mutation subtending $b$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (identical for all $x$, given the uniform labeling distribution).
Let $I_k$ denote the $k$th intercoalescent time interval, i.e.\ $I_n = (0, T_n),\ I_{n-1} = (T_n, T_{n-1}),\ \dots,\ I_2 = (T_3, T_2)$.
Let rv $M_{\delta x, b, k}$ denote the event that the mutation $M_{\delta x, b}$ occurred during $I_k$.
For small $\delta x$ and finite $\mu(t)$ we have
\begin{align*}
\mathbbm{P}(M_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(M_{\delta x, b, k}\mid T)\\
&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{t\in I_k}\mu(t)dt + O(\delta x)\right),
\end{align*}
where
\begin{equation}
\label{eqn:p}
p_{n,k}(b) = \frac{\binom{n-b-1}{k-2}}{\binom{n-1}{k-1}}
\end{equation}
is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9), and the quantity in parenthesis is the probability that a mutation arose during the $k$th intercoalescent interval in a small genomic interval of size $\delta x$.
Marginalizing $T$ gives
\begin{align*}
\mathbbm{P}(M_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right].
\end{align*}
For small $\delta x$, each such interval contains zero or one mutations, so the expected number of mutations subtending $b$ haplotypes (i.e.\ the $b$th component of the SFS) is
\[
\xi_b = \int d\mathbbm{P}(M_{dx, b}) = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
\]
Using the explicit bounds of the coalescent intervals gives
\begin{align}
\label{eqn:xi}
\xi_b &= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=2}^{n-1} k p_{n,k}(b) \mathbbm{E}_{T_{k+1}}\left[\int_0^{T_{k+1}}\mu(t)dt\right]\nonumber\\
&= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=3}^{n} (k-1) p_{n,k-1}(b) \mathbbm{E}_{T_{k}}\left[\int_0^{T_k}\mu(t)dt\right]\nonumber\\
&= \sum_{k=2}^n B_{b,k} \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right],
\end{align}
where
\begin{equation}
\label{eqn:B}
B_{b,k}\equiv
\begin{cases}
k p_{n,k}(b),& k=2\\
k p_{n,k}(b) - (k-1) p_{n,k-1}(b),& k > 2
\end{cases}
\end{equation}

@@

We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
\begin{equation}
\label{eqn:xivec}
\boldsymbol\xi = C \int_0^\infty \boldsymbol r(t)\mu(t)dt,
\end{equation}
where the $(n-1)$ by $(n-1)$ matrix
\[
C_{b,j} \equiv \sum_{k=2}^j B_{b,k} A_{k,j}
\]
is constant wrt $\mu$ \emph{and} $\eta$.

We next develop an efficient recursive procedure for computing $C$.
Using \ref{eqn:B}
\begin{align*}
C_{b,j} &= \sum_{k=2}^j k p_{n,k}(b) A_{k,j} - \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}\\
&= W_{b,j} - W'_{b,j}
\end{align*}
Where $W$ is defined recursively in \cite{Polanski2003-ll} (eqs 13-15), and we define
\begin{equation}
\label{eqn:W'}
W'_{b,j} = \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
\end{equation}
\cite{Polanski2003-ll} (eqn 11) show that $A$ can be expressed as
\[
A_{k,j} = \frac{n! (n-1)!}{(j+n-1)! (n-j)!} \frac{(2 j-1)}{j (j-1)} \frac{(j+k-2)!}{ (k-1)! (k-2)! (j-k)! }(-1)^{j-k},
\]
so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \ref{eqn:W'} is a definite sum over geometric terms.
Zeilberger's algorithm \citep{petkovvsek1996b, paule1995mathematica} can thus be used to yield the following second-order recursion in $j$:
\begin{align*}
&W'_{b,2} = 0\\
&W'_{b,3} = \frac{(-1)^{j-3} (2 j-1) (b-1)! (j+1)! n! (-b+n-1)! \binom{n-2}{b-1}}{(j-1) j (j-3)! (n-j)! (j+n-1)!}\\
&W'_{b,j+2} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W'_{b,j}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W'_{b,j+1}\right)
\end{align*}

This completes the discrete mathematics needed to frame mutation rate inference as a linear inverse problem.
We next consider eqn \ref{eqn:xivec} as a problem in numerical analysis.

@@

\section*{Results}\label{sec:results}

@@

\section*{Discussion}\label{sec:discussion}

>>>>

\section{Introduction}\label{sec:intro}

@@

\section{Model}\label{sec:model}

\subsection{The expected SFS given demographic and mutation rate histories}\label{sec:model:xi}

@@

Let rv $\mathcal{E}_{\delta x, b}$ denote the event that a mutation subtending $b$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (identical for all $x$, given the uniform labeling distribution).
Let $I_k$ denote the $k$th intercoalescent time interval, i.e.\ $I_n = (0, T_n),\ I_{n-1} = (T_n, T_{n-1}),\ \dots,\ I_2 = (T_3, T_2)$.
Let rv $\mathcal{E}_{\delta x, b, k}$ denote the event that the mutation $\mathcal{E}_{\delta x, b}$ occurred during $I_k$.
For small $\delta x$ and finite $\mu(t)$ we have
\begin{align*}
\mathbbm{P}(\mathcal{E}_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(\mathcal{E}_{\delta x, b, k}\mid T)\\
&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{t\in I_k}\mu(t)dt + O(\delta x)\right),
\end{align*}
where
\begin{equation}
\label{eqn:p}
p_{n,k}(b) = \frac{\binom{n-b-1}{k-2}}{\binom{n-1}{k-1}}
\end{equation}
is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9), and the quantity in parenthesis is the probability that a mutation arose during the $k$th intercoalescent interval in a small genomic interval of size $\delta x$.
Marginalizing $T$ gives
\begin{align*}
\mathbbm{P}(\mathcal{E}_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right].
\end{align*}
For small $\delta x$, each such interval contains zero or one mutations, so the expected number of mutations subtending $b$ haplotypes (i.e.\ the $b$th component of the SFS) is
\[
\xi_b = \int d\mathbbm{P}(\mathcal{E}_{dx, b}) = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
\]
Using the explicit bounds of the coalescent intervals gives
\begin{align}
\label{eqn:xi}
\xi_b &= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=2}^{n-1} k p_{n,k}(b) \mathbbm{E}_{T_{k+1}}\left[\int_0^{T_{k+1}}\mu(t)dt\right]\nonumber\\
&= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=3}^{n} (k-1) p_{n,k-1}(b) \mathbbm{E}_{T_{k}}\left[\int_0^{T_k}\mu(t)dt\right]\nonumber\\
&= \sum_{k=2}^n B_{b,k} \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right],
\end{align}
where
\begin{equation}
\label{eqn:B}
B_{b,k}\equiv
\begin{cases}
k p_{n,k}(b),& k=2\\
k p_{n,k}(b) - (k-1) p_{n,k-1}(b),& k > 2
\end{cases}
\end{equation}

@@

We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
\begin{equation}
\label{eqn:xivec}
\boldsymbol\xi = C \boldsymbol d,
\end{equation}
where the $(n-1)$ by $(n-1)$ matrix
\[
C_{b,j} \equiv \sum_{k=2}^j B_{b,k} A_{k,j}
\]
is constant wrt $\mu$ \emph{and} $\eta$, and
\begin{equation}
\label{eqn:d}
d_j \equiv \int_0^\infty r_j(t)\mu(t)dt
\end{equation}
has linear (nonlinear) dependence on $\mu$ ($\eta$).

\subsection{Computing the elements of $C$}\label{sec:model:C}

We next develop an efficient recursive procedure for computing $C$.
Using \ref{eqn:B}
\begin{align*}
C_{b,j} &= \sum_{k=2}^j k p_{n,k}(b) A_{k,j} - \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}\\
&= W_{b,j} - W'_{b,j}
\end{align*}
Where $W$ is defined recursively in \cite{Polanski2003-ll} (eqns 13-15), and we define $W'$ by
\begin{equation}
\label{eqn:W'}
W'_{b,j} = \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
\end{equation}
\cite{Polanski2003-ll} (eqn 11) show that $A$ can be expressed as
\[
A_{k,j} = \frac{n! (n-1)!}{(j+n-1)! (n-j)!} \frac{(2 j-1)}{j (j-1)} \frac{(j+k-2)!}{ (k-1)! (k-2)! (j-k)! }(-1)^{j-k},
\]
so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \ref{eqn:W'} is a definite sum over geometric terms.
Zeilberger's algorithm \citep{petkovvsek1996b, paule1995mathematica} can thus be used to yield the following second-order recursion in $j$:
\begin{align*}
&W'_{b,2} = 0\\
&W'_{b,3} = \frac{20 (n-2)}{(n+1)(n+2)}\\
&W'_{b,j+2} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W'_{b,j}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W'_{b,j+1}\right)
\end{align*}
This completes the framing of mutation rate inference as a linear inverse problem.

\subsection{Finite dimensional parameterization of $\eta(t)$ and $\mu(t)$}\label{sec:model:pcsws}

Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \ref{eqn:d} to give
\begin{equation}
\label{eqn:d2}
d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)dt,
\end{equation}
where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
We consider piecewise constant $\eta$ and $\mu$ on $\ell$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{\ell-1}, t_\ell)$ where $0=t_0 < t_1 < \dots < t_{\ell-1} < t_\ell=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
Let $(y_1,\dots,y_\ell)$ denote the constant population size $eta(t)$ during each epoch, and let $(z_1,\dots,z_\ell)$ denote the constant mutation rate $mu(t)$ during each epoch.
Let $x_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=2,\dots,t_\ell$, and $x_0\equiv 1$.
With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
\begin{equation}
\label{eqn:d3}
\boldsymbol d = M(\boldsymbol y) \boldsymbol z
\end{equation}
where
\begin{equation}
\label{eqn:M}
M(\boldsymbol y) \equiv
\begin{bmatrix}
1 &             &        &                       \\
  & \frac{1}{3} &        &                       \\
  &             & \ddots &                       \\
  &             &        & \frac{1}{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1       & x_1                & \hdots & \prod_{i=1}^{\ell-1}x_i               \\
1       & x_1^3              & \hdots & \prod_{i=1}^{\ell-1}x_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & x_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{\ell-1}x_i^{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
y_1  &      &        &             &       \\
-y_1 & y_2  &        &             &       \\
     & -y_2 & y_3    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -y_{\ell-1} & y_\ell
\end{bmatrix}.
\end{equation}
Note that the $(n-1)$ by $\ell$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $x_l$ are nonlinear functions of $\boldsymbol y$.

Combining \ref{eqn:d3} with \ref{eqn:xivec} gives the discretized inverse problem
\begin{equation}
\label{eqn:xivec2}
\boldsymbol\xi = C M(\boldsymbol y) \boldsymbol z.
\end{equation}

@@

\section{Results}\label{sec:results}

@@

\section{Discussion}\label{sec:discussion}
